On the specific recursive structure of iterated parametricity

### 2025-11-11 paper/paper.tex
Reworking text + fixing layout and compilation

\begin{align*}
  \framep[n-2][p] & : \U                                                                                    \\
  \restrf[n-1][p] & : \Pi q \leq n+p.\, \framep[n-1][p] \rightarrow \framep[n-2][p]                         \\
  \cohf[n][p]     & : \Pi qr\epsilon\omega.\, \restrf[n-1][p][\epsilon][q] \circ \restrf[n][p][\omega][r] = \\
                  & \;\;\;\;\restrf[n-1][p][\omega][r] \circ \restrf[n][p][\epsilon][q+1]                   \\
  \ldots          & \;\;\;\;\texttt{fields for painting, restrf, cohp}
\end{align*}
where $\framep[n][p]$, $\framep[n-1][p]$ and $\restrf[n][p]$ are now appropriate definitions.

Turning $\framep[n][p]$, $\framep[n-1][p]$ and $\restrf[n][p]$ into definitions is however tricky: the definition of $\framep[n-1][p]$, which is by induction on $p$ (for all $n$), depends on $\restrf[n-1][p]$ whose type depends on $\framep[n][p-1]$. This means that $\framep[n-1][p]$ and the type of $\restrf[n-1][p]$ have to be mutually defined.

The well-foundedness of the construction can be justified by the observation that:

Defining $\framep[n][p]$ for $p \leq n$ requires the definition of $\painting[n-1][p-1]$, thus also $\framep[n-1][p-1]$ by typing dependency, as well as $\restrf[n][p-1]$, where $\restrf[n][p-1]$ in turn depends in $\framep[n][p-1]$ for its typing.

Moreover, $\framep[n][p]$ depends on $\framep[n][p-1]$, so these dependencies are between the blocks of $\framep[n][k]$ for all $k\leq p$, as well as $\framep[n-1][k]$, $\painting[n-1][k]$ and $\restrf[n][k]$ for all $k\leq p-1$, which we respectively abbreviate as $\framep[n][0,...,p]$, $\framep[n-1][0,...,p-1]$, $\painting[n-1][0,...,p-1]$ and $\restrf[n][0,...,p-1]$.

Also, $\restrf[n][p-1]$ depends on $n$.

Schematically, using dotted arrows for dependencies in the type and plain arrows for dependencies in the definition, this gives the following dependencies for $p\leq n$ fixed:
\begin{center}
  \fontsize{7.2}{9}\selectfont
  \begin{tikzcd}
    \framep[n][0,...,p] \arrow[d] \arrow[ddr]\\
    \painting[n-1][0,...,p-1] \arrow[d, dotted] \\
    \framep[n-1][0,...,p-1] &
    \restrf[n][0,...,p-1] \arrow[l, dotted] \arrow[uul, dotted, "\mbox{$[0,...,p-1]$}"', near start, shift right=2mm, hook']\\
  \end{tikzcd}
\end{center}

In particular, the mutual dependency on one side of $\framep[n][0,...,p-1]$ within the type of $\restrf[n][0,...,p-1]$ and on the other side of $\restrf[n][0,...,p-1]$ within the definition of $\framep[n][0,...,p]$ requires to mutually define the type of $\restrf[n][0,...,p-1]$ together with $\framep[n][p]$ (or $\framep[n][0,...,p]$) as a function from $\restrf[n][0,...,p-1]$.

To talk about sequences of $\framep$, $\painting$ or $\restrf$, we need to specify their types. The type of $\restrf$ will be defined mutually with $\framep$ of the same level, but the type of $\framep$, $\painting$ can be defined in advance, the second one being however dependent on the first one. In general, components of such sequences will have a type dependent on the previous components of the sequence, as it is the case for $\restrf$. So, the type of a sequence of length $p$ will be expressed as a $n$-iterated $\Sigma$-type. In the case of $\framep$ and $\painting$, there will be however no dependency, so an ordinary product can be used. If $\framep[n][0,...,p]$ is such a sequence of length $p+1$, we write $\framep[n][0,...,p-1]$ for its immediate subsequence of length $p$ and $\framep[n][p]$ for its last component.

@@

\subsection{Specifying frames and paintings}

We can now define the types of sequences of $\framep$ and $\painting$ as follows:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \frametype[n][0,...,p]    &                                              & :      & \U                                                                                           \\
    \frametype[n][]           &                                              & \defeq & \unittype                                                                                    \\
    \frametype[n][0,...,p]    &                                              & \defeq & \frametype[n+1][0,...,p-1] \times \U                                                         \\
    \\
    \paintingtype[n][0,...,p] & (\framep[n][0,...,p]:\frametype[n][0,...,p]) & :      & \U                                                                                           \\
    \paintingtype[n][]        & \unitpoint                                   & \defeq & \unittype                                                                                    \\
    \paintingtype[n][0,...,p] & (\framep[n][0,...,p])                        & \defeq & \paintingtype[n+1][0,...,p-1](\framep[n+1][0,...,p-1]) \times (\framep[n][p] \rightarrow \U) \\
  \end{array}
\end{equation*}

@@

Frames are defined mutually with the types of frame restrictions,
using auxiliary definitions for the types and definitions of layers:

@@

    & \defeq & \Pi q \leq n. \Pi \epsilon. \Pi d: \framep[n+1][p]

@@

>>>>

\begin{align*}
  \framep[n][p] & : \U & \mbox{for all $p\leq n$}                                                                                      \\
  \painting[n][p] & : \framep[n][p] \rightarrow \U & \mbox{for all $p\leq n$}                                                          \\
  \restrf[n][p] & : \Pi q \leq n+p.\, \framep[n+1][p] \rightarrow \framep[n][p] & \mbox{for all $p\leq n$}                         \\
  \restrp[n][p] & : \Pi q \leq n+p.\, \Pi d:\framep[n+1][p]\, \painting[n+1][p](d) \rightarrow \painting[n][p](\restrf[n][p](d)) & \mbox{for all $p\leq n$} \\
  \cohf[n][p]   & : \Pi qr\epsilon\omega.\, \restrf[n][p][\epsilon][q] \circ \restrf[n+1][p][\omega][r] = \\
                  & \;\;\;\;\restrf[n][p][\omega][r] \circ \restrf[n+1][p][\epsilon][q+1] & \mbox{for all $p\leq n$}                   \\
  \cohp[n][p]     & \Pi qr\epsilon\omega.\,\Pi d:\framep[n+2][p].\, \restrp[n][p][q][\epsilon](\restrf[n+1][p][r][\omega](d)) \circ \restrp[n+1][p][r][\omega](d) = \\
                  & \;\;\;\;\restrp[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon](d)) \circ \restrp[n+1][p][q+1][\epsilon](d) & \mbox{for all $p\leq n$}                 \\
\end{align*}
where $\framep[n+2][p]$, $\framep[n+1][p]$ and $\restrf[n+1][p]$, as
well as $\painting[n+2][p]$, $\painting[n+1][p]$ and $\restrp[n+1][p]$
are now appropriate definitions.

Turning $\framep[n+2][p]$, $\framep[n+1][p]$ and $\restrf[n+1][p]$ into definitions involves however circularities. To start with, the definition of $\framep[n+1][p]$, which is by induction on $p$, depends on $\restrf[n][p-1]$ whose type depends in turn on $\framep[n+1][p-1]$. This means that $\framep[n+1][p]$ and the type of $\restrf[n][p]$ have to be mutually defined.

The definition of $\framep[n+1][p]$ also depends, via $\layer[n][p-1]$, on $\painting[n][p-1]$. Both via $\painting[n][p-1]$ and $\restrf[n][p-1]$, it depends also on $\framep[n][p-1]$ by typing dependency. This requires to assume $\framep[n][p-1]$ and $\painting[n][p-1]$ known in advance.
Moreover, the definition of $\framep[n+1][p]$ depends on $\framep[n+1][p-1]$, so all of $\framep[n][k]$, $\painting[n][k]$ and $\restrf[n][k]$ for $k< p$ need to be known, which we respectively abbreviate as $\framep[n][0,...,p-1]$, $\painting[n][0,...,p-1]$ and $\restrf[n][p-1]$.
Schematically, using dotted arrows for dependencies in the type and plain arrows for dependencies in the definition, this gives the following dependencies for $p\leq n$ fixed:
\begin{center}
  \fontsize{7.2}{9}\selectfont
  \begin{tikzcd}
    \framep[n+1][0,...,p] \arrow[d] \arrow[ddr]\\
    \painting[n][0,...,p-1] \arrow[d, dotted] \\
    \framep[n][0,...,p-1] &
    \restrf[n][0,...,p-1] \arrow[l, dotted] \arrow[uul, dotted, "\mbox{$[0,...,p-1]$}"', near start, shift right=2mm, hook']\\
  \end{tikzcd}
\end{center}

In particular, the mutual dependency on one side of $\framep[n][0,...,p-1]$ within the type of $\restrf[n][0,...,p-1]$ and on the other side of $\restrf[n][0,...,p-1]$ within the definition of $\framep[n][0,...,p]$ requires to mutually define the type of $\restrf[n][0,...,p-1]$ together with $\framep[n][p]$ (or $\framep[n][0,...,p]$) as a \emph{function} from $\restrf[n][0,...,p-1]$.

So far, this means that we are able to specify a dependent structure of the form:
\begin{align*}
  \framep[n][p] & : \U & \mbox{for all $p\leq n$}                                                                                      \\
  \painting[n][p] & : \framep[n][p] \rightarrow \U & \mbox{for all $p\leq n$}                                                          \\
  \restrf[n][p] & : \Pi q \leq n+p.\, \framep[n+1][p]
    \left(\begin{array}{l}
      \framep[n][0,...,p-1]   \\
      \painting[n][0,...,p-1] \\
      \restrf[n][0,...,p-1]   \\
    \end{array}\right)
  \rightarrow \framep[n][p] & \mbox{for all $p\leq n$}                         \\
\end{align*}
which can appropriately be reformulated using sequences for appropriate definitions of $\frametype[n][0,...,p]$, $\paintingtype[n][0,...,p]$ and $\restrftype[n][0,...,p]$, since the dependencies of $\restrf[n][p]$ into $\restrf[n][0,...,p-1]$ makes it a very dependent function~\cite{Hickey96}:
\begin{align*}
  \framep[n][0,...,p] & : \frametype[n][0,...,p]                                                                                       \\
  \painting[n][0,...,p] & : \paintingtype[n][0,...,p](\framep[n][0,...,p])                                                         \\
  \restrf[n][0,...,p] & : \restrftype[n][0,...,p]
    \left(\begin{array}{l}
      \framep[n][0,...,p-1]   \\
      \painting[n][0,...,p-1] \\
      \restrf[n][0,...,p-1]   \\
    \end{array}\right)\\
\end{align*}

In general, the types of sequences such as $\restrf[n][0,...,p]$ will be nested since components, such as $\restrf[n][p]$, may depend on previous components of the sequence. So, the type of a sequence of length $p$ will be expressed as a $p$-iterated $\Sigma$-type. In the case of $\framep$ and $\painting$, there will be however no dependency, so an ordinary product can be used. If $\framep[n][0,...,p]$ is such a sequence of length $p+1$, we write $\framep[n][0,...,p-1]$ for its immediate subsequence of length $p$ and $\framep[n][p]$ for its last component, and similarly for all such sequences.

@@

\subsection{Specifying frames}

Assuming $\framep[n][0,...,p]$ only requires to define the type of frames, which are just finite products of $\U$:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \frametype[n][0,...,p]    &                                              & :      & \U                                                                                           \\
    \frametype[n][]           &                                              & \defeq & \unittype                                                                                    \\
    \frametype[n][0,...,p]    &                                              & \defeq & \frametype[n+1][0,...,p-1] \times \U                                                         \\
  \end{array}
\end{equation*}

\subsection{Specifying paintings}

The types $\paintingtype[n][0,...,p]$ of $\painting[n][0,...,p]$
are type families over frames, so $\framep[n][0,...,p]$ needs to be assumed
beforehand:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \paintingtype[n][0,...,p] & (\framep[n][0,...,p]:\frametype[n][0,...,p]) & :      & \U                                                                                           \\
    \paintingtype[n][]        & \unitpoint                                   & \defeq & \unittype                                                                                    \\
    \paintingtype[n][0,...,p] & (\framep[n][0,...,p])                        & \defeq & \paintingtype[n+1][0,...,p-1](\framep[n+1][0,...,p-1]) \times (\framep[n][p] \rightarrow \U) \\
  \end{array}
\end{equation*}

@@

Assuming $\framep[n][0,...,p-1]$ and $\painting[n][0,...,p-1]$,
we can mutually define the types $\restrftype[n][0,...,p-1]$ of frame
restrictions together with $\framep[n+1][0,...p]$ as a function
from $\restrf[n][0,...,p-1]$, using auxiliary definitions for the
types and definitions of layers:

@@

    & \defeq &
    \Pi q \leq n. \Pi \epsilon. \Pi d: \framep[n+1][p].\\

@@

    \end{array}\right.\\
